Bird beak's elaboration and innovation

### 2022-07-20 Analysis/03.1-Elaboration_innovation_analysis[statistics].Rmd
Update 03.1-Elaboration_innovation_analysis[statistics].Rmd

```{r, eval = TRUE, echo = FALSE}
## Load the data for all birds
load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order.rda")
## Select the data
shapespace_allbirds <- shapespace_allbirds_lvl_superorder_order[[1]]$space
## Clean garbage
rm(shapespace_allbirds_lvl_superorder_order)

@@

## Create the subset for all birds
all_birds_shapespace <- MCMCglmm.subsets(n = 500, # Just run it on 500 matrices for now
      data          = shapespace_allbirds,
      posteriors    = shapespace_allbirds_posteriors,
      group         = MCMCglmm.levels(shapespace_allbirds_posteriors)[c(super_lvl, order_lvl, phylo_lvl)],
      rename.groups = c(levels(shapespace_allbirds$level1)[-1], levels(shapespace_allbirds$level2)[-1], "phylogeny")) #[-1] is because the first level is ""
## Clean garbage
rm(shapespace_allbirds_posteriors)

@@

## Create the subset for all birds
passeriformes_shapespace <- MCMCglmm.subsets(n = 500, # Just run it on 500 matrices for now
      data          = shapespace_passeriformes,
      posteriors    = shapespace_passeriformes_posteriors,
      group         = MCMCglmm.levels(shapespace_passeriformes_posteriors)[c(suborder_lvl, family_lvl, phylo_lvl)],
      rename.groups = c(levels(shapespace_passeriformes$level1)[-1], levels(shapespace_passeriformes$level2)[-1], "phylogeny")) #[-1] is because the first level is ""
## Clean garbage
rm(shapespace_passeriformes_posteriors)
```

```{r}
## Creating the nested levels subgroups
make.nested.subset <- function(group, data, group.data) {
    ## Get the subsets within the group
    level2_groups <- as.character(unique(group.data$level2[group.data$level1 == group]))
    if(any(empty <- level2_groups == "")) {
        level2_groups <- level2_groups[-which(empty)]
    }
    if(length(level2_groups) < 1) {
        return(NULL)
    }
    ## Create a subset
    return(get.subsets(data, subsets = c(level2_groups, group)))
}

@@

```{r, eval = TRUE, echo = FALSE}
load("../Data/Processed/ellipse_stats.rda")
## Ploting fun
ploting.stats <- function(results, col) {

    col <- col[match(names(size.subsets(results[[1]])), names(col))]

    layout_matrix <- layout(matrix(1:3, nrow = 1), widths = c(4.2, 3, 3), heights = n.subsets(results[[1]])/2)
    par(mar = c(5, 10, 2, 1) + 0.1)
    plot(results[[1]], horizontal = TRUE, las = 1, xlab = "angle", ylab = "", col = col)
    par(mar = c(5, 1, 2, 1) + 0.1)
    plot(results[[2]], horizontal = TRUE, las = 1, xlab = "disalignment", ylab = "", yaxt = "n", col = col)
    par(mar = c(5, 1, 2, 1) + 0.1)
    plot(results[[3]], horizontal = TRUE, las = 1, xlab = "distance between centroids", ylab = "", yaxt = "n", col = col)
}

## Ploting fun
table.stats <- function(results, class) {
    ## Table results
    table <- cbind(summary(results[[1]])[, -c(1:3)], summary(results[[2]])[, -c(1:3)], summary(results[[3]])[, -c(1:3)])
    rownames(table) <- gsub(":phy", "", as.vector(summary(results[[1]])[,1]))
    colnames(table)[c(1,6,11)] <- c("angle", "disalignment", "distance")
    if(!missing(class)) {
        table <- cbind(class = class, table)
    }
    return(table)
}

all_results <- rbind(table.stats(ellipse_stats$super_orders_stats, class = "superorder"),
                     table.stats(ellipse_stats$orders_stats, class = "orders"),
                     table.stats(ellipse_stats$suborder_stats, class = "suborder"),
                     table.stats(ellipse_stats$family_stats, class = "family"))
knitr::kable(all_results)
```

```{r, eval = TRUE, echo = FALSE}
load("../Data/Processed/tip_colours_super_orders.rda")
ploting.stats(ellipse_stats$super_orders_stats, col = tip_colours_super_orders)
```

```{r, eval = TRUE, echo = FALSE}
load("../Data/Processed/tip_colours_orders.rda")
ploting.stats(ellipse_stats$orders_stats, col = tip_colours_orders)
```

```{r, eval = TRUE, echo = FALSE}
load("../Data/Processed/tip_colours_passeriformes_suborder.rda")
ploting.stats(ellipse_stats$suborder_stats, col = tip_colours_passeriformes_suborder)
```

```{r, eval = TRUE, echo = FALSE}
load("../Data/Processed/tip_colours_passeriformes_family.rda")
ploting.stats(ellipse_stats$family_stats, col = tip_colours_passeriformes_family)
```

>>>>

```{r, eval = FALSE, echo = FALSE}
## Load the data for all birds
load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order.rda")
## Select the data
shapespace_allbirds <- shapespace_allbirds_lvl_superorder_order[[1]]$space
## Clean garbage
rm(shapespace_allbirds_lvl_superorder_order)

@@

## Create the subset for all birds
all_birds_shapespace <- MCMCglmm.subsets(n = 4000, # Just run it on 500 matrices for now
      data          = shapespace_allbirds,
      posteriors    = shapespace_allbirds_posteriors,
      group         = MCMCglmm.levels(shapespace_allbirds_posteriors)[c(super_lvl, order_lvl, phylo_lvl)],
      rename.groups = c(levels(shapespace_allbirds$level1)[-1], levels(shapespace_allbirds$level2)[-1], "phylogeny")) #[-1] is because the first level is ""
## Clean garbage
rm(shapespace_allbirds_posteriors)

@@

## Create the subset for all birds
passeriformes_shapespace <- MCMCglmm.subsets( # Just run it on 500 matrices for now
      data          = shapespace_passeriformes,
      posteriors    = shapespace_passeriformes_posteriors,
      group         = MCMCglmm.levels(shapespace_passeriformes_posteriors)[c(suborder_lvl, family_lvl, phylo_lvl)],
      rename.groups = c(levels(shapespace_passeriformes$level1)[-1], levels(shapespace_passeriformes$level2)[-1], "Passeriformes")) #[-1] is because the first level is ""
## Clean garbage
rm(shapespace_passeriformes_posteriors)
```

```{r, eval = FALSE, echo = FALSE}
## Creating the nested levels subgroups
make.nested.subset <- function(group, data, group.data) {
    ## Get the subsets within the group
    level2_groups <- as.character(unique(group.data$level2[group.data$level1 == group]))
    if(any(empty <- level2_groups == "")) {
        level2_groups <- level2_groups[-which(empty)]
    }
    if(length(level2_groups) < 1) {
        return(NULL)
    }
    ## Create a subset
    return(get.subsets(data, subsets = c(level2_groups, group)))
}

@@

And here are the full results! Here's what's in each column:

 * **Group**: the name of the group (a random term of the model), the focal VCV
 * **Comparison**: to which group are the VCV compared (the group is either one or two levels higher)
 * **n**: number of species
 * **distance**: the distance between the focal and the comparison group's centers
 * **ellipse sd**: the standard deviation of the angles in the the focal VCV distribution
 * **disalignment**: the disalignment of the group compared to the focal group (0 = aligned), with the 95% CI
 * **orthogonality**: the orthogonality amount between both groups (0 = parallel, 1 = orthogonal), with the 95% CI
 * **Post. prob.**: the posterior probability of the orthogonality in the focal group being different from the comparison one (with the 95% CI estimated from the randomised posterior probabilities)

```{r, eval = TRUE, echo = FALSE}
load("../Data/Processed/ellipses_stats.rda")

## Ploting fun
table.stats <- function(results) {
    ## Table results

    ## Names, comparison and size
    header <- summary(results$distances)[,c(1,2), drop = FALSE]
    head_1 <- as.data.frame(do.call(rbind, strsplit(header[,1], split = ":")))
    header <- cbind(head_1, "n" = header[, 2])
    colnames(header)[c(1,2)] <- c("Group", "Comparison")

    ## Adding the statistics
    table <- cbind(header, "distance" = summary(results$distances)[,4])
    table <- cbind(table, "ellipse sd" = results$sd[-length(results$sd)])
    table <- cbind(table, summary(results$alignments, quantiles = c(95))[, -c(1:3)])
    colnames(table)[ncol(table)-2] <- "disalignment"
    table <- cbind(table, summary(results$angles, quantiles = c(95))[, -c(1:3)])
    colnames(table)[ncol(table)-2] <- "orthogonality"
    table <- cbind(table, results$test)
    colnames(table)[ncol(table)-2] <- "Post. prob."
    rownames(table) <- NULL
    return(table)
}

## Sorting all results
all_results <- do.call(rbind, lapply(ellipses_stats[-12], table.stats)) ## Droping the Meliphagoidea (single group)
rownames(all_results) <- NULL

write.csv(all_results, "../Data/Processed/ellipses.stats.csv")
knitr::kable(all_results)
```

<!-- length(which(all_results[which(all_results[,"Post. prob."] >= 0.95), ]$orthogonality > 0.5)) -->

```{r, eval = TRUE, echo = FALSE, fig.height = 20, fig.width = 8}
# pdf("../Manuscript/Figures/orthogonality.results.pdf", width = 6, height = 15)
source("../Functions/ploting.utilities.R")
## Plot the orthogonality into context
orthogonality <- all_results[, c(1:2, 9:14)]

## Group per level
super_orders <- orthogonality[c(1:8), ]
orders <- orthogonality[c(36:60), ]
orders_n2 <- orthogonality[c(9:35), ]
sub_orders <- orthogonality[c(61:66), ]
families <- orthogonality[c(90:110), ]
families_n2 <- orthogonality[c(67:89), ]

## Sort each level by orthogonality score
ortho.sort <- function(data) {
    return(data[order(data$orthogonality, decreasing = TRUE), , drop = FALSE])
}
## Plotting table (with the sorted orthogonality)
plot_table <- rbind(ortho.sort(super_orders),
                    ortho.sort(orders),
                    ortho.sort(sub_orders),
                    ortho.sort(families))
colours <- c(rep(gg.color.hue(4)[1], nrow(super_orders)),
             rep(gg.color.hue(4)[2], nrow(orders)),
             rep(gg.color.hue(4)[3], nrow(sub_orders)),
             rep(gg.color.hue(4)[4], nrow(families)))

## Adding starts here and there
get.signif <- function(prob) {
    if(prob >= 0.99) {
        return("***")
    } else {
        if(prob >= 0.95) {
            return("**")
        } else {
            if(prob >= 0.90) {
                return("*")
            } else {
                if(prob >= 0.80) {
                    return(".")
                } else {
                    return("")
                }
            }
        }
    }
}
add.stars <- function(data, counter, shift = 0, total, ...) {
    text(data[3] + 0.03, (total+1)-counter + shift + 0.2, labels = get.signif(data[6]), cex = 1, ...)   
}
add.line <- function(data, counter, shift = 0, total, ...) {
    lines(matrix(c(data[4:5], rep((total+1)-counter + shift, 2)), 2, 2), ...)
    points(data[3], (total+1)-counter + shift, pch = 19, ...)
}

## Plot frame
par(mar = c(5, 9, 4, 2))
plot(NULL, yaxt = "n", xaxt = "n", ylab = "", xlab = "orthogonality", ylim = c(1,nrow(plot_table)), xlim = c(0,1))
abline(v = 0.5, lwd = 0.5, col = "grey")
abline(v = 0.75, lty = 2, col = "grey", lwd = 0.5)
axis(2, at = nrow(plot_table):1, labels = plot_table$Group, las = 2)
axis(1, at = c(0, 0.5, 1), labels = c(0, 0.5, 1))
## Adding the results

for(i in 1:nrow(plot_table)) {
    add.line(plot_table[i, ], counter = i, col = colours[i], total = nrow(plot_table))
    add.stars(plot_table[i, ], counter = i, col = colours[i], total = nrow(plot_table))
}

## Adding the n2 results
orders_n2_plot <- orders_n2[match(orders[order(orders$orthogonality, decreasing = TRUE), ]$Group, orders_n2$Group), ]
families_n2_plot <- families_n2[match(families[order(families$orthogonality, decreasing = TRUE), ]$Group, families_n2$Group), ]
for(i in 1:nrow(orders_n2_plot)) {
    add.line(orders_n2_plot[i, ], shift = -1/3, counter = i+nrow(super_orders), col = adjustcolor(colours[i+nrow(super_orders)], alpha.f = 1/3), total = nrow(plot_table))
    add.stars(orders_n2_plot[i, ], shift = -1/3, counter = i+nrow(super_orders), col = adjustcolor(colours[i+nrow(super_orders)], alpha.f = 1/3), total = nrow(plot_table))
}
for(i in 1:nrow(orders_n2)) {
    add.line(families_n2_plot[i, ], shift = -1/3, counter = i+(nrow(plot_table)-nrow(families)), col = adjustcolor(colours[i+(nrow(plot_table)-nrow(families))], alpha.f = 0.5), total = nrow(plot_table))
    add.line(families_n2_plot[i, ], shift = -1/3, counter = i+(nrow(plot_table)-nrow(families)), col = adjustcolor(colours[i+(nrow(plot_table)-nrow(families))], alpha.f = 0.5), total = nrow(plot_table))
}

legend("topleft", legend = c("Super orders (vs. phylogeny)", "Orders (vs. super orders)", "Orders (vs. phylogeny)", "Suborders (vs. passeriformes)", "Families (vs. suborders)", "Families (vs. passeriformes)"), col = c(gg.color.hue(4)[1], gg.color.hue(4)[2], adjustcolor(gg.color.hue(4)[2], alpha.f = 1/3), gg.color.hue(4)[3], gg.color.hue(4)[4], adjustcolor(gg.color.hue(4)[4], alpha.f = 1/3)), pch = 19, lty = 1, cex = 0.8)
# dev.off()
```

Distribution of the VCV ellipses' orthogonality for all the groups compared to the ellipse of their parent clade (or n-2). 
For the posterior probabilities: . >= 0.8; * >= 0.9; ** >= 0.95; *** >= 0.99.

```{r, eval = FALSE, echo = FALSE}
load("../Data/Processed/tip_colours_super_orders.rda")
ploting.stats(ellipse_stats$super_orders_stats, col = tip_colours_super_orders)
```

```{r, eval = FALSE, echo = FALSE}
load("../Data/Processed/tip_colours_orders.rda")
ploting.stats(ellipse_stats$orders_stats, col = tip_colours_orders)
```

```{r, eval = FALSE, echo = FALSE}
load("../Data/Processed/tip_colours_passeriformes_suborder.rda")
ploting.stats(ellipse_stats$suborder_stats, col = tip_colours_passeriformes_suborder)
```

```{r, eval = FALSE, echo = FALSE}
load("../Data/Processed/tip_colours_passeriformes_family.rda")
ploting.stats(ellipse_stats$family_stats, col = tip_colours_passeriformes_family)
```
